---
title: "ch_1_exercises"
output: html_document
date: "2022-10-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#If install of package loader is required then isntall it
if (!require(pacman)) {
  install.packages("pacman")
}

# Changes any scientific number notation to normal
options(scipen = 999) 

#Load extra packages
pacman::p_load(data.table, tidyverse, lubridate, patchwork, 
               MASS, magrittr, readr, plm, GGally)

```

1. Create a vector called x and place values 10 to 1000. Create a ğ‘¦ vector that takes the square root of the log of numbers in ğ‘¥. Plot ğ‘¥ and ğ‘¦

```{r}

x <- 10:1000
y <- sqrt(log(x))

ggplot() +
  geom_point(aes(x, y))

```

2. In the R ğğšğ­ğšğ¬ğğ­ package there is a data set called ğšğ¢ğ«ğªğ®ğšğ¥ğ¢ğ­ğ². In Python, load the ğšğ¢ğ«ğªğ®ğšğ¥ğ¢ğ­ğ².ğœğ¬ğ¯ data file from the GitHub repository

```{r}

airquality <- data.table(readr::read_csv("../../data/airquality.csv", show_col_types = FALSE))

head(airquality, 4)

```

3. Mean, median and 75th percentile values of ğ“ğğ¦ğ©

```{r}

mean(airquality$Temp)
median(airquality$Temp)
quantile(airquality$Temp, 0.75)

```

4. Create a subset dataframe called ğ¬1 which has data for all days in June ### not explained in text

```{r}

s1 <- airquality[Month == 6]
head(s1, 5)

```

4. Draw a plot from ğ¬1 of ğ“ğğ¦ğ© and ğ’ğ¨ğ¥ğšğ«.ğ‘

```{r}

ggplot(s1) +
  geom_point(aes(Temp, Solar.R))

```

5. What is the median value of ğ’ğ¨ğ¥ğšğ«.ğ‘ in the ğ¬1 dataframe?

```{r}

median(s1$Solar.R)

```

6. In R, load the MASS package. There is a data set called Cars93. In Python, load the 93ğœğšğ«ğ¬.ğœğ¬ğ¯ data file

```{r}

cars <- data.table(MASS::Cars93)
head(cars, 20)

```

7. Mean, median, and 25th percentile values of ğŒğğ† on high way

```{r}

mean(cars$MPG.highway)
median(cars$MPG.highway)
quantile(cars$MPG.highway, 0.25)

```

8. Create a subset dataframe called ğ©1 which has data for all Non-4WD cars that weigh over 3500

9. Draw a plot from ğ©1 of price and horsepower

```{r}

p1 <- cars[(DriveTrain != "4WD") & (Weight >= 3500)]

ggplot(p1) +
  geom_point(aes(Price, Horsepower))

```

10. What is the median luggage room in the ğ©1 dataframe?

```{r}

median(na.omit(p1$Luggage.room))

```

11. In R, install the package ğ©ğ¥ğ¦, load the library and load the data using the command ğğšğ­ğš(ğŒğšğ¥ğğ¬). In Python, load ğŒğšğ¥ğğ¬.ğœğ¬ğ¯ from the repository.

```{r}

males <- data.table(readr::read_csv("../../data/Males.csv", show_col_types = FALSE))

head(males, 5)

```

12. Create a frequency table between industry and residence. ### not explained in text Where do people in manufacturing mostly live and by what proportion? What proportion of people in the south are in Transportation? What is the most popular industry for people living in rural_area?

```{r}

males[, .N, by = c("industry", "residence")][order(industry, residence)]

```

