---
title: "Chapter-2-ExerciseSolutions"
output: html_document
date: "2022-10-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#If install of package loader is required then isntall it
if (!require(pacman)) {
  install.packages("pacman")
}

# Changes any scientific number notation to normal
options(scipen = 999) 

#Load extra packages
pacman::p_load(data.table, tidyverse, lubridate, patchwork, 
               MASS, magrittr, readr, plm, GGally)

```

```{r}

gapminder = data.table(readr::read_csv('../../data/gapminder.csv', show_col_types = FALSE))

head(gapminder , 4)

```

1. Calculate the average life expectancy for each year.

```{r}

gapminder[, .(lifeExp = mean(lifeExp)), by = "year"]

```

2. Plot year versus life expectancy and draw a horizontal line at the average life expectancy value. Color the points based on the continent

```{r}

ggplot(gapminder) +
  geom_point(aes(year, lifeExp, color = continent)) +
  geom_hline(aes(yintercept = mean(lifeExp)), color = "blue")

```

3. Calculate the total population for each continent for each year. Create an appropriate plot to visualize the data

```{r}

gapminder[, .(pop = sum(pop)), by = "year"]

```

```{r}

ggplot(gapminder) +
  geom_col(aes(year, pop / 1000000000)) +
  labs(title = "Population per year",
       y = "population (billions)")

```

4. Plot GDP per capita for each year for African countries. Add appropriate labels to indicate the country name.

```{r}

af = gapminder[continent == "Africa"]

head(af, 5)

```

```{r, fig.height = 8}

ggplot(af, aes(year, gdpPercap)) +
  geom_line() +
  facet_wrap(vars(country))

```

```{r}

usedcars = readr::read_csv('../../data/UsedCars2017.csv',
                           show_col_types = FALSE)

head(usedcars, 5)

```

5. Create a plot of ğ€ğ ğ and ğŒğ¢ğ¥ğğšğ ğ. What conclusions can you draw from the plot?

```{r}

ggplot(usedcars) +
  geom_point(aes(Age, Mileage))

```

6. Explore the data by creating various plots and document any insights from visualization

```{r}

summary(usedcars)

```

```{r}

GGally::ggpairs(usedcars, progress = FALSE)

```

